---
title: "Chapter 7"
output: html_notebook
---

```{r}
library(tidyverse)
library(nycflights13)
library(ggstance)
library(lvplot)
library(ggbeeswarm)
library(hexbin)
library(modelr)
```

# Visualizing Distributions
```{r}
# Categorical Variables
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))

diamonds %>%
  count(cut)

# Continuous variables
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)

diamonds %>%
  count(cut_width(carat, 0.5))

# Zooming in
smaller <- diamonds %>%
  filter(carat < 3)

ggplot(data = smaller) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.1)

smaller %>%
  count(cut_width(carat, 0.1))

# Overlaying data from multiple histograms
ggplot(data = smaller, mapping = aes(x = carat, color = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

# Typical values/Unusual values
```{r}
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

ggplot(data = faithful, mapping = aes(x = eruptions)) +
  geom_histogram(binwidth = 0.25)

ggplot(diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

# Zoom in
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-5, 65))

(unusual <- diamonds %>%
    filter(y < 3 | y > 20) %>%
    select(price, x, y, z) %>%
    arrange(y))
```

# Exercises
```{r}
# 1
(diamonds_xyz <- diamonds %>%
   mutate(id = row_number()) %>%
   select(x, y, z, id) %>%
   gather(variable, value, -id) %>%
   ggplot(mapping = aes(x = value)) +
    geom_density() +
    geom_rug() +
    facet_grid(variable ~ .)
 )
summarize(diamonds, mean(x >= y), mean(x > z), mean(y > z))
```

# Exercises
```{r}
# 2
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram(binwidth = 10)

ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram(binwidth = 10) +
  coord_cartesian(xlim = c(0, 5000))

ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram(binwidth = 50) +
  coord_cartesian(xlim = c(0, 5000))
```

# Exercises (cont.)
```{r}
# 3
car1 <- dim(diamonds %>%
              filter(carat == 0.99))[1]

car2 <- dim(diamonds %>%
              filter(carat == 1))[1]
abs(car1 - car2)

(diamonds %>%
    filter((carat >= 0.99 & carat <= 1)) %>%
    count(carat))

(diamonds %>%
    filter((carat >= 0.9 & carat <= 1.1)) %>%
    count(carat))
```

# Exercises (cont.)
```{r}
# 4
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat)) +
  coord_cartesian(xlim = c(0, 3), ylim = c(0, 8500))

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat)) +
  xlim(0, 3) +
  ylim(0, 8500)

# Try zooming halfway (whole bar disappears)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat)) +
  xlim(0, 3) +
  ylim(0, 4250)
```

# Missing Values
```{r}
# Find the outliers (can also refer to the graph in a previous section)
(diamonds %>%
   filter((y < 2 | y > 15)))

# drop all strange values -- not recommended because you might just run out of data
(diamonds_clean <- diamonds %>%
   filter(between(y, 3, 20)))

# can replace the unusual values with missing values
(diamonds_clean <- diamonds %>%
    mutate(y = ifelse(y < 3 | y > 20, NA, y)))

# When using ggplot, removing values will yield a warning
ggplot(data = diamonds_clean, mapping = aes(x = x, y = y)) +
  geom_point()

# Suppress the warning
ggplot(data = diamonds_clean, mapping = aes(x = x, y = y)) +
  geom_point(na.rm = TRUE)
```

```{r}
(nycflights13::flights %>%
   mutate(
     cancelled = is.na(dep_time),
     sched_hour = sched_dep_time %/% 100,
     sched_min = sched_dep_time %% 100,
     sched_dep_time = sched_hour + sched_min / 60
   ) %>%
   ggplot(mapping = aes(sched_dep_time)) +
    geom_freqpoly(mapping = aes(color = cancelled), binwidth = 1/4))
```

# Exercises
```{r}
(flts1 <- flights %>%
   mutate(
     dep_delay = ifelse(rowSums(is.na(flights)) > 0, NA, dep_delay)
   ))

ggplot(data = flts1, mapping = aes(x = dep_delay)) +
  geom_histogram(binwidth = 10)

(flts2 <- flights %>%
    mutate(
      carrier = ifelse(rowSums(is.na(flights)) > 0, NA, carrier)
    ))

ggplot(data = flts2, mapping = aes(x = carrier)) +
  geom_bar()

mean(flts1$dep_delay)
mean(flts1$dep_delay, na.rm = TRUE)
sum(flts1$dep_delay)
sum(flts1$dep_delay, na.rm = TRUE)
```

# Covariation
```{r}
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_freqpoly(mapping = aes(color = cut), binwidth = 500)

ggplot(diamonds) +
  geom_bar(mapping = aes(x = cut))

# Swap count for density
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) +
  geom_freqpoly(mapping = aes(color = cut), binwidth = 500)

# Using boxplots
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()

# Dealing with unordered factors
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()

ggplot(data = mpg, mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  geom_boxplot()

# Flip plots to deal with long variable names
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip()
```

# Exercises
```{r}
# 1
(flts <- flights %>%
   mutate(
     cancelled = is.na(dep_time) | is.na(arr_time),
     sched_hour = dep_time %/% 100,
     sched_min = dep_time %% 100,
     sched_dep_time = sched_hour + sched_min / 60
   ))

ggplot(data = flts, mapping = aes(x = sched_dep_time, y = ..density..)) +
  geom_freqpoly(mapping = aes(color = cancelled))

ggplot(data = flts, mapping = aes(x = cancelled, y = sched_dep_time)) +
  geom_boxplot()
```

# Exercises (cont.)
```{r}
# 2
# See how carat relates to price
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point()

# Try a binned boxplot
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))

# See how carat relates to cut
ggplot(data = diamonds, mapping = aes(x = reorder(cut, carat, FUN = median), y = carat)) +
  geom_boxplot()
```

# Exercises (cont.)
```{r}
# 3
# Using coord_flip()
ggplot(data = diamonds, mapping = aes(x = reorder(cut, carat, FUN = median), y = carat)) +
  geom_boxplot() +
  coord_flip()

# Using ggstance
ggplot(data = diamonds, mapping = aes(y = reorder(cut, carat, FUN = median), x = carat)) +
  geom_boxploth()

# 4
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_lv()
```

# Exercises (cont.)
```{r}
# geom_freqpoly()
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) +
  geom_freqpoly(mapping = aes(color = cut), binwidth = 500)

# facetted geom_histogram()
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram() +
  facet_wrap(~ cut, ncol = 1, scales = "free_y")

ggplot(data = diamonds, mapping = aes(x = price, y = cut)) +
  geom_violinh()
```

# Exercises
```{r}
ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median),
                                 y = hwy))

ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median),
                                 y = hwy),
                   method = "tukey")

ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median),
                                 y = hwy),
                   method = "tukeyDense")

ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median),
                                 y = hwy),
                   method = "frowney")

ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(x = reorder(class, hwy, FUN = median),
                                 y = hwy),
                   method = "smiley")

ggplot(data = mpg) +
  geom_beeswarm(mapping = aes(x = reorder(class, hwy, FUN = median),
                              y = hwy))
```

# Two categorical variables
```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))

diamonds %>%
  count(color, cut)

diamonds %>%
  count(color, cut) %>%
  ggplot(mapping = aes(x = color, y = cut)) +
  geom_tile(mapping = aes(fill = n))
```

# Exercises
```{r}
# 1
# Use a proportion rather than a count
(diamonds %>%
   count(color, cut) %>%
   group_by(color) %>%
   mutate(prop = n / sum(n)) %>%
   ggplot(mapping = aes(x = color, y = cut)) +
     geom_tile(mapping = aes(fill = prop)))

(diamonds %>%
    count(color, cut) %>%
    group_by(cut) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(mapping = aes(x = color, y = cut)) +
      geom_tile(mapping = aes(fill = prop)))
```

# Exercises (cont.)
```{r}
(flights_grpd <- flights %>%
    group_by(month, dest) %>%
    summarize(dep_delay = mean(dep_delay, na.rm = TRUE)))

ggplot(data = flights_grpd, mapping = aes(x = factor(month), y = dest, fill = dep_delay)) +
  geom_tile() +
  labs(x = "Month", y = "Destination", fill = "Departure Delay")

(flights_grpd <- flights %>%
    group_by(month, dest) %>%
    summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
    group_by(dest) %>%
    # At least one flight each month
    filter(n() == 12) %>%
    ungroup() %>%
    mutate(dest = reorder(dest, dep_delay)))

ggplot(data = flights_grpd, mapping = aes(x = factor(month), y = dest, fill = dep_delay)) +
  geom_tile()
```

# Two continuous variables
```{r}
# Covariation example
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))

# Correct for overplotting by using the alpha aesthetic
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price), alpha = 1 / 50)

# 2d binning
ggplot(data = diamonds) +
  geom_bin2d(mapping = aes(x = carat, y = price))

ggplot(data = diamonds) +
  geom_hex(mapping = aes(x = carat, y = price))

# Using binning
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)), varwidth = TRUE)

# Try to put approx. same number of points in each bin
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))
```

# Exercises
```{r}
# 1
ggplot(data = diamonds,
       mapping = aes(color = cut_number(carat, 5), x = price)) +
  geom_freqpoly() +
  labs(y = "Carat", x = "Price")

ggplot(data = diamonds,
       mapping = aes(color = cut_width(carat, 0.5, boundary = 0), x = price)) +
  geom_freqpoly() +
  labs(y = "Carat", x = "Price")

ggplot(diamonds, aes(x = carat, y = cut_number(price, 10))) +
  geom_boxploth() +
  labs(y = "Price", x = "Carat")

ggplot(diamonds, aes(x = carat, y = cut_width(price, 2000, boundary = 0))) +
  geom_boxploth(varwidth = TRUE) +
  labs(y = "Price", x = "Carat")

```

# Exercises (cont.)
```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d() +
  facet_wrap(~ cut) +
  labs(x = "Carat", y = "Price")

ggplot(diamonds, aes(x = price, y = cut_number(carat, 5), color = cut)) +
  geom_boxploth() +
  labs(y = "Carat", x = "Price")

ggplot(diamonds, aes(color = cut_number(carat, 5), x = price, y = cut)) +
  geom_boxploth(varwidth = TRUE)

ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

# Patterns and Models
```{r}
ggplot(data = faithful) +
  geom_point(mapping = aes(x = eruptions, y = waiting))

# use a model to predict price from carat and then compute residuals
mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>%
  add_residuals(mod) %>%
  mutate(resid = exp(resid))

ggplot(data = diamonds2) +
  geom_point(mapping = aes(x = carat, y = resid, alpha = 1 / 50))

# Now that the relationship between carat and price is removed, take another look at cut and price
ggplot(data = diamonds2) +
  geom_boxploth(mapping = aes(x = resid, y = reorder(cut, resid, FUN = median)))

```

# ggplot2 calls
```{r}
ggplot(faithful, aes(eruptions)) +
  geom_freqpoly(binwidth = 0.25)

diamonds %>%
  count(cut, clarity) %>%
  ggplot(aes(clarity, cut, fill = n)) +
    geom_tile()
```

