---
title: "Chapter 5"
output: html_notebook
---


```{r}
library(nycflights13)
library(tidyverse)
```

# nycflights13
```{r}
flights
```

# Filter Rows
```{r}
# filter by flights on the 1st of january
(jan1 <- filter(flights, month == 1, day == 1))
```

# Comparisons
```{r}
sqrt(2) ^ 2 == 2
1 / 49 * 49 == 1
near(sqrt(2) ^ 2, 2)
near(1 / 49 * 49, 1)

```

# Logical operators
```{r}
(nov_dec <- filter(flights, month == 11 | month == 12))

# using shorthand
(nov_dec <- filter(flights, month %in% c(11, 12)))
# using demorgan's
(delay_range1 <- filter(flights, !(arr_delay > 120 | dep_delay > 120)))
```

# Missing Values
```{r}
NA > 5
10 == NA
NA == NA
x <- NA
is.na(x)
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```

# Exercises
```{r}
# 1
filter(flights, arr_delay >= 120)
filter(flights, dest == "IAH" | dest == "HOU")
filter(flights, carrier == "UA" | carrier == "AA" | carrier == "DL")
filter(flights, month %in% c(7, 8, 9))
filter(flights, dep_delay <= 0, arr_delay > 120)
filter(flights, dep_delay >= 60, dep_delay - arr_delay > 30)
filter(flights, between(dep_time, 0, 600) | dep_time == 2400)
filter(flights, is.na(dep_time))
```

# Arrange Rows
```{r}
arrange(flights, year, month, day)
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)

# Missing values are always sorted at the end
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```

# Exercises
```{r}
# 1
arrange(flights, desc(is.na(dep_time)), dep_time)
# 2
# Most delayed
arrange(flights, desc(dep_delay))
# Earliest
arrange(flights, dep_delay)
# 3
# total air time
arrange(flights, air_time)
# average air time
arrange(flights, distance / air_time * 60)
# 4
# interpretation 1
arrange(flights, desc(distance))
arrange(flights, distance)
# interpretation 2
arrange(flights, desc(air_time))
arrange(flights, air_time)
```

# Select Columns
```{r}
select(flights, year, month, day)
# Select range of columns
select(flights, year:day)
# Exclude columns
select(flights, -(year:day))
rename(flights, tail_num = tailnum)
# Move some variables to the start of the dataframe
select(flights, time_hour, air_time, everything())
```

# Exercises
```{r}
# 1
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, "dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, starts_with("dep_"), starts_with("arr_"))
select(flights, 4, 6, 7, 9)
select(flights, matches("^(dep|arr)_(delay|time)$"))
of_interest <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, one_of(of_interest))
# 2
select(flights, dep_time, dep_time)

# 4
select(flights, contains("TIME"))
select(flights, contains("TIME", ignore.case = FALSE))
```

# Add new variables
```{r}
(flights_sml <- select(
  flights,
  year:day,
  ends_with("delay"),
  distance,
  air_time
))

# add some columns
(mutate(
  flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
))

# Use new columns for an opperation
(mutate(
  flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
))

# Only keep the new variables
(transmute(
  flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
))

# Using other functions
(transmute(
  flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
))

# Lag and Lead
(x <- 1:10)
lag(x)
lead(x)

# Running aggregates
cumsum(x)
cummean(x)

# Ranking
y <- c(2, 3, 3, NA, 4, 7)
min_rank(y)
min_rank(desc(y))
row_number(y)
dense_rank(y)
cume_dist(y)
```

# Exercises
```{r}
# 1
# Midnight is represented as 2400 -> 24 * 60 = 1440 minutes after midnight (need to account for this)
time_to_mins <- function(t) {
  return((t %/% 100 * 60 + t %% 100) %% 1440)
}

(flight_times <- mutate(
  flights,
  dep_time_mins = time_to_mins(dep_time),
  sched_dep_time_mins = time_to_mins(sched_dep_time)
))

select(flight_times, dep_time, sched_dep_time, dep_time_mins, sched_dep_time_mins)

# 2
# Grab rows of interest
(flight_durs <- select(flights, arr_time, dep_time, air_time, origin, dest))
# Convert them all to minutes (except air time, which already is)
(flight_durs <- mutate(
  flight_durs,
  arr_time_mins = time_to_mins(arr_time),
  dep_time_mins = time_to_mins(dep_time)
))
# Rearrange columns
flight_durs <- select(flight_durs, origin, dest, air_time, everything())
# Calculate air_time_diff
(flight_durs <- mutate(flight_durs, air_time_diff = arr_time_mins - dep_time_mins))
# Check if it worked. Are there negatives?
arrange(flight_durs, air_time_diff)
# There are, so flights that go overnight are incorrect. Also, timezones could cause problems

# 3
# SHould be difference between dep time and scheduled dep time
flights_deptime <- 
  mutate(
    flights,
    dep_time_mins = time_to_mins(dep_time),
    sched_dep_time_mins = time_to_mins(sched_dep_time),
    dep_delay_mins = dep_time_mins - sched_dep_time_mins
  )
# Have same issue that overnight delays appear as <= 0
```

# Exercises (cont.)
```{r}
# 4
(most_delayed <- select(
  mutate(
    flights,
    dep_delay_rank = min_rank(desc(dep_delay))
  ),
  dep_delay_rank,
  dep_delay,
  everything()
  )
 )
arrange(most_delayed, dep_delay)
# Only grab the 50 most delayed (there are no repeats in the top 10)
(top_50 <- arrange(
  filter(most_delayed, dep_delay_rank <= 50),
  dep_delay_rank)
)

# 5
1:3 + 1:10
# keeps cycling 1:3 until reach 10
```

# Grouped Summaries
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
by_day <- group_by(flights, year, month, day)
summarize(by_day, delay = mean(dep_delay, na.rm = TRUE))

# Group flights by destination
by_dest <- group_by(flights, dest)
# Summarize distance, avg. delay, number of flights
(delays <- summarize(
  by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
))
# Honolulu is quite far away from all others
arrange(delays, desc(dist))
# Ignore destinations that rarely occur
arrange(delays, count)
# Filter to remove noise (I.E. Honolulu and destinations with a very small number of flights)
(delays <- filter(delays, count > 20, dest != "HNL"))

# Visualize the data
ggplot(data = delays, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

# Do the same thing using pipes
# Note: x %>% f(y) == f(x, y) and x %>% f(y) %>% g(z) == g(f(x, y), z)
(delays <- flights %>%
  group_by(dest) %>%
  summarize(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dest != "HNL") %>%
  arrange(desc(delay))
)

# Reason for using na.rm
flights %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay))
# Fixed
flights %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay, na.rm = TRUE))

# Remove cancelled flights from the data set
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay))
```

# Grouped Summaries (cont.)
```{r}
# Find individual planes (identified by tail num) and their average delays
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) +
  geom_freqpoly(binwidth = 10)

# Try with number of flights vs avg delay
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(
    delay = mean(arr_delay),
    flight_cnt = n()
  )

# we see that variation decreases as sample size increases
ggplot(data = delays, mapping = aes(x = flight_cnt, y = delay)) +
  geom_point(alpha = 1/10)

# Filter out groups with smallest number of observation
delays %>%
  filter(flight_cnt > 25) %>%
  ggplot(mapping = aes(x = flight_cnt, y = delay)) +
    geom_point(alpha = 1/10)
```

# Baseball Example
```{r}
# Convert to tibble
batting <- as_tibble(Lahman::Batting)

(batters <- batting %>%
  group_by(playerID) %>%
  summarize(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  ))

batters %>%
  filter(ab > 100) %>%
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() +
    geom_smooth(se=FALSE)

# Find people with highest averages (these will likely be people with only a few at bats)
batters %>%
  arrange(desc(ba))
```

# Useful summary functions
```{r}
# Logical subsetting
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    avg_delay = mean(arr_delay),
    avg_pos_delay = mean(arr_delay[arr_delay > 0]) # avg. delay amongst positive delays
  )

# Measures of spread
not_cancelled %>%
  group_by(dest) %>%
  summarize(distance_sd = sd(distance)) %>%
  arrange(desc(distance_sd))

# Measures of rank
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    first = min(dep_time),
    last = max(dep_time)
  )

# Measures of position
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    first = first(dep_time),
    last = last(dep_time)
  )

# Filtering
not_cancelled %>%
  group_by(year, month, day) %>%
  mutate(r = min_rank(desc(dep_time))) %>%
  select(year:day, r, everything()) %>%
  filter(r %in% range(r))

# Counts
not_cancelled %>%
  group_by(dest) %>%
  summarize(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))

not_cancelled %>%
  count(dest)

not_cancelled %>%
  count(tailnum, wt = distance)

# How many flights left before 5am each day
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(n_early = sum(dep_time < 500))

# What proportion of flights are delayed by more than an hour?
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(lg_delay = mean(arr_delay > 60))
```

# Grouping by multiple variables
```{r}
(daily <- group_by(flights, year, month, day))
(per_day <- summarize(daily, flight_cnt = n()))
(per_month <- summarize(per_day, flight_cnt = sum(flight_cnt)))
(per_year <- summarize(per_month, flight_cnt = sum(flight_cnt)))

# Ungroup
daily %>%
  ungroup() %>%
  summarize(flights = n()) # all flights
```

# Exercises
```{r}
# 2
not_cancelled %>%
  group_by(dest) %>%
  # length() finds number of observations in each group
  summarize(n = length(dest))

# can accomplish the same thing using n()
not_cancelled %>%
  group_by(dest) %>%
  summarize(n = n())

# or can use group_by() and tally()
not_cancelled %>%
  group_by(dest) %>%
  tally()

# Weighted count
not_cancelled %>%
  group_by(tailnum) %>%
  summarize(n = sum(distance))

# Or can use group_by() and tally()
not_cancelled %>%
  group_by(tailnum) %>%
  tally(distance)

# 4
(cancelled_delayed <- flights %>%
  mutate(cancelled = (is.na(dep_delay) | is.na(arr_delay) | is.na(dep_time))) %>%
  group_by(year, month, day) %>%
  summarize(
    flight_cnt = n(),
    prop_cancelled = mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE)
  )
)

ggplot(cancelled_delayed, aes(x = avg_dep_delay, y = prop_cancelled)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

# Exercises (cont.)
```{r}
carr <- flights %>%
  group_by(carrier)
# Find worst carrier for departures
(carr %>%
   summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
   arrange(desc(avg_dep_delay))
 )
# Find worst carrier for arrivals
(carr %>%
   summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
   arrange(desc(avg_arr_delay))
) 

(ap_fl <- flights %>%
  group_by(dest) %>%
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE))
)

(car_fl <- flights %>%
  group_by(carrier, dest) %>%
  summarize(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
  ) %>%
  filter(avg_arr_delay >= 0 | avg_arr_delay < 0) # gets rid of NaNs
)
```

# Grouped Mutates
```{r}
# Find worst members of each group
(flights_sml %>%
   group_by(year, month, day) %>%
   filter(rank(desc(arr_delay)) < 10))

# Find all groups bigger than a threshold
(popular_dests <- flights %>%
  group_by(dest) %>%
  filter(n() > 365)
)

# Standardize to compute per group metrics
popular_dests %>%
  filter(arr_delay > 0) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>%
  select(year:day, dest, arr_delay, prop_delay)
```

# Exercises
```{r}
# 2
(flights %>%
   group_by(tailnum) %>%
   summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
   select(tailnum, avg_arr_delay) %>%
   filter(min_rank(desc(avg_arr_delay)) <= 10) %>%
   arrange(desc(avg_arr_delay))
)

# 3
(flights %>%
    group_by(hour) %>%
    summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
    filter(min_rank(avg_arr_delay) <= 10) %>%
    arrange(avg_arr_delay)
)

# 4
(flights %>%
    filter(!is.na(arr_delay), arr_delay > 0) %>%
    group_by(dest) %>%
    # mutate(arr_delay_total = sum(arr_delay))
    mutate(
      arr_delay_total = sum(arr_delay),
      arr_delay_prop = arr_delay / arr_delay_total
    )
)

# 5
(lagged_delays <- flights %>%
  arrange(origin, year, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))
)

# Plot the relationship
lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarize(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
# Summarize relationship
lagged_delays %>%
  summarize(delay_diff = mean(dep_delay - dep_delay_lag, na.rm = TRUE))
```

# Exercises (cont.)
```{r}
# 6
# Find unusual observations by standardizing the data
standardized_flights <- flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest, origin) %>%
  mutate(air_time_mean = mean(air_time),
         air_time_sd = sd(air_time),
         n = n()) %>%
  ungroup() %>%
  mutate(air_time_standard = (air_time - air_time_mean) / air_time_sd)

ggplot(standardized_flights, aes(x = air_time_standard)) +
  geom_density()

ggplot(standardized_flights, aes(x = air_time_mean, y = air_time_standard)) +
  geom_point()
```

# Exercises (cont.)
```{r}
standardized_flights %>%
  arrange(air_time_standard) %>%
  mutate(date = str_c(year, "-", month, "-", day, " ", dep_time),
         flightnum = str_c(carrier, flight)) %>%
  select(date, flightnum, origin, dest, distance, air_time, air_time_mean, air_time_standard) %>%
  head(10) %>%
  print(width = 100)

# 7
# Step 1: find all airports serviced by 2 or more carriers
(dest_2 <- flights %>%
    # Keep only uinique carrier/dest pairs
    select(dest, carrier) %>%
    group_by(dest, carrier) %>%
    filter(row_number() == 1) %>%
    # count carriers by destination
    group_by(dest) %>%
    mutate(n_carrier = n_distinct(carrier)) %>%
    filter(n_carrier >= 2)
)

# Step 2: rank carriers by the number of destinations that they service
(carriers_by_dest <- dest_2 %>%
    group_by(carrier) %>%
    summarize(n_dest = n()) %>%
    arrange(desc(n_dest)))
head(carriers_by_dest)

# 8
(flights %>%
    select(year:day, tailnum, dep_delay, arr_delay) %>%
    group_by(tailnum) %>%
    arrange(tailnum, year, month, day) %>%
    filter(!is.na(dep_delay), !is.na(arr_delay)) %>%
    mutate(delay_gt1hr = dep_delay > 60) %>%
    mutate(before_delay = cumsum(delay_gt1hr)) %>%
    filter(before_delay < 1) %>%
    count(sort = TRUE)
)
```
